---
title: "Causal Inference: Synthetic Control Estimation"
output: html_document
date: "2026-02-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This notebook estimates causal treatment effects using an elastic-net synthetic
control method. The workflow:

1. Load treatment and control data produced by the similarity search.
2. Reshape to wide format and split into pre/post-treatment periods.
3. Fit elastic-net models to construct synthetic controls for each unit.
4. Compute gaps (residuals) and RMSPE ratios for inference.
5. Visualize the synthetic control vs. treatment unit trajectories.

---

## 1. Setup & Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(glmnet)
library(ggplot2)
library(ggthemes)
library(MLmetrics)
library(tidyverse)
library(dplyr)
```

---

## 2. Load Data & Configuration

```{r load-data}
# NOTE: Update this path to point to your working directory
setwd("/Users/phoebely/Documents/Projects/Wellspring")

control_final <- read.csv("control_final_v2.csv")
treat_final   <- read.csv("treat_final_v2.csv")

# Average treatment group values by year
# Dynamically select all numeric columns for averaging
average_treat <- treat_final %>%
  group_by(year) %>%
  summarise(across(where(is.numeric), mean, na.rm = TRUE)) %>%
  ungroup()

```

```{r config}
# Change TREAT_YEAR
TREAT_YEAR <- 2020            # Year treatment begins

# Derive year range and counts from the data itself
ALL_YEARS  <- sort(unique(c(control_final$year, treat_final$year)))
YEAR_MIN   <- min(ALL_YEARS)
YEAR_MAX   <- max(ALL_YEARS)
N_PRE      <- sum(ALL_YEARS < TREAT_YEAR)
N_POST     <- sum(ALL_YEARS >= TREAT_YEAR)
```

---

## 3. Reshape to Wide Format

Stack the control and treatment panels, then pivot so each column is a unit's
AWI time series.

```{r reshape}
sample_TC <- rbind(control_final, treat_final)

wide_TC <- sample_TC %>%
  select(UID, year, awi) %>%
  pivot_wider(names_from = UID, values_from = awi, names_prefix = "uid_")

# Identify unit columns (everything except "year")
unit_cols <- setdiff(names(wide_TC), "year")
N_UNITS   <- length(unit_cols)

# Split into pre- and post-treatment matrices (unit columns only)
wide_pre_df  <- wide_TC %>% filter(year <  TREAT_YEAR)
wide_post_df <- wide_TC %>% filter(year >= TREAT_YEAR)

wide_pre  <- as.matrix(wide_pre_df[, unit_cols])
wide_post <- as.matrix(wide_post_df[, unit_cols])

cat("Pre-treatment matrix: ", dim(wide_pre),  "\n")
cat("Post-treatment matrix:", dim(wide_post), "\n")
```

---

## 4. Elastic-Net Synthetic Control Loop

For each unit, fit an elastic-net regression (alpha = 0.5) on the pre-treatment
period using all other units as donors. Weights are constrained to [0, 1] with
no intercept.



```{r elastic-net-loop, warning=FALSE}
# Remove units with any missing values in pre or post periods
complete_pre  <- complete.cases(t(wide_pre))
complete_post <- complete.cases(t(wide_post))
complete_units <- complete_pre & complete_post

wide_pre  <- wide_pre[, complete_units]
wide_post <- wide_post[, complete_units]
unit_cols <- unit_cols[complete_units]
N_UNITS   <- length(unit_cols)

cat("Units after dropping NAs:", N_UNITS, "\n")

# Pre-allocate results: one row per unit, columns = pre gaps + post gaps + 4 stats
N_COLS     <- N_PRE + N_POST + 4  # gaps + RMSPE_post + RMSPE_pre + ratio + pct_change
output_rmspe <- matrix(NA, nrow = N_UNITS, ncol = N_COLS)
rownames(output_rmspe) <- unit_cols

for (i in seq_along(unit_cols)) {

  target_col <- unit_cols[i]
  donor_cols <- setdiff(unit_cols, target_col)

  # Fit elastic-net: predict target unit from all donor units in pre-treatment period
  res <- glmnet(
    x = wide_pre[, donor_cols],
    y = wide_pre[, target_col],
    family       = "gaussian",
    lower.limits = 0,
    upper.limits = 1,
    intercept    = FALSE,
    standardize  = FALSE,
    alpha        = 0.5
  )

  lambda_min <- min(res$lambda)

  # Predict synthetic control for pre and post periods
  pre_pred  <- predict(res, newx = wide_pre[, donor_cols],  s = lambda_min)
  post_pred <- predict(res, newx = wide_post[, donor_cols], s = lambda_min)

  # Store pre-treatment gaps (actual - synthetic)
  output_rmspe[i, seq_len(N_PRE)] <- wide_pre[, target_col] - pre_pred

  # Store post-treatment gaps
  output_rmspe[i, N_PRE + seq_len(N_POST)] <- wide_post[, target_col] - post_pred

  # RMSPE: post, pre, ratio, and percent change
  gap_col <- N_PRE + N_POST
  output_rmspe[i, gap_col + 1] <- RMSPE(post_pred, wide_post[, target_col])
  output_rmspe[i, gap_col + 2] <- RMSPE(pre_pred,  wide_pre[, target_col])
  output_rmspe[i, gap_col + 3] <- output_rmspe[i, gap_col + 1] / output_rmspe[i, gap_col + 2]
  output_rmspe[i, gap_col + 4] <- ((post_pred[N_POST] - post_pred[1]) / abs(post_pred[1])) * 100
}
```

---

## 5. Placebo Gap Plot

Plot the gap (actual − synthetic) for every unit. The treatment unit's gap
is highlighted in bold.

```{r gap-plot, fig.width=10, fig.height=6}
TIME_POINTS  <- ALL_YEARS
GAP_COLS     <- seq_len(N_PRE + N_POST)

treat_uid    <- paste0("uid_", unique(treat_final$UID))
treat_row    <- which(unit_cols %in% treat_uid)
control_rows <- setdiff(seq_along(unit_cols), treat_row)

graph <- output_rmspe[control_rows, GAP_COLS]

matplot(t(graph), type = "l", xlab = "Year", ylab = "Gap in AWI", xaxt = "n")

# Single averaged treatment line
treat_avg <- colMeans(output_rmspe[treat_row, GAP_COLS], na.rm = TRUE)
matlines(treat_avg, type = "l", col = 1, lwd = 5)

axis(2)
axis(side = 1, at = seq_len(ncol(graph)), labels = TIME_POINTS)

```

---

## 6. Build Synthetic Control vs. Treatment Comparison

Extract the treatment unit's effect estimates and construct confidence bands
using RMSPE.

```{r}
# Average across all treatment units
gap_col    <- N_PRE + N_POST
treat_gaps <- colMeans(output_rmspe[treat_row, , drop = FALSE], na.rm = TRUE)

rmspe_post <- treat_gaps[gap_col + 1]
rmspe_pre  <- treat_gaps[gap_col + 2]

# Year-by-year effect (gap) for averaged treatment
effect_data <- data.frame(
  year   = ALL_YEARS,
  effect = treat_gaps[GAP_COLS]
)

# Observed AWI for treatment group
result_data <- average_treat %>% select(year, awi)

merged_df <- effect_data %>%
  left_join(result_data, by = "year") %>%
  mutate(
    RMSPE_post        = rmspe_post,
    RMSPE_pre         = rmspe_pre,
    synthetic_control = awi - effect,
    RMSPE_post_upper  = awi + abs(awi) * RMSPE_post,
    RMSPE_post_lower  = awi - abs(awi) * RMSPE_post,
    RMSPE_pre_upper   = awi + abs(awi) * RMSPE_pre,
    RMSPE_pre_lower   = awi - abs(awi) * RMSPE_pre
  )
```


```{r summary-stats}
outcome_col <- 'awi'

summary_stats <- average_treat %>%
  group_by(year) %>%
  summarise(
    mean     = signif(mean(.data[[outcome_col]]), 7),
    p5_awi   = quantile(.data[[outcome_col]], 0.05),
    p25_awi  = quantile(.data[[outcome_col]], 0.25),
    p75_awi  = quantile(.data[[outcome_col]], 0.75),
    p95_awi  = quantile(.data[[outcome_col]], 0.95)
  )

comb <- merged_df %>% left_join(summary_stats, by = "year")
```

---

## 7. Final Visualization

```{r synth-plot, fig.width=10, fig.height=6}
GRAPH_TITLE <- "Synthetic Control vs. Treatment Units"

ggplot(comb, aes(x = year)) +
  # Confidence bands
  geom_ribbon(aes(ymin = RMSPE_post_lower, ymax = RMSPE_post_upper),
              fill = "lightgrey", alpha = 0.5) +
  geom_ribbon(aes(ymin = RMSPE_pre_lower, ymax = RMSPE_pre_upper),
              fill = "darkgrey", alpha = 0.5) +
  # Treatment and synthetic control lines
  geom_line(aes(y = mean, color = "Treatment Units"), linewidth = 1) +
  geom_line(aes(y = synthetic_control, color = "Synthetic Control"), linewidth = 1) +
  # Treatment year marker
  geom_vline(xintercept = TREAT_YEAR, linetype = "dotted", color = "black") +
  # Labels and theme
  labs(x = "Year", y = "AWI", title = GRAPH_TITLE) +
  scale_color_manual(
    values = c("Synthetic Control" = "blue", "Treatment Units" = "red"),
    name   = NULL
  ) +
  scale_fill_manual(
    values = c("lightgrey", "darkgrey"),
    labels = c("Outer CI", "Inner CI")
  ) +
  theme_bw() +
  theme(
    legend.position      = "top",
    legend.justification = "center",
    legend.key           = element_blank()
  ) +
  guides(fill = guide_legend(title = NULL))
```
